# Package {#package}

`r newthought("As with many other things in R,")` the threshold of complexity
for writing your own package has recently lowered dramatically. If you are writing
some of your own functions in R to use for your research, and you've become comfortable
with writing functions, you should try putting them in an R package. You don't have
to share this package through publishing it on CRAN or another repository---instead, it's 
fine to start by just writing packages for your own use, or for your research group's 
private use. However, once you start writing packages, you will see how straightforward
it is, and you may want to take the extra steps to prepare your package for CRAN and 
submit it.

An R package is just "a directory of files which extend R" (https://cran.r-project.org/doc/manuals/r-release/R-exts.html).
You should consider writing your own package anytime you find yourself cutting and pasting
functions to different directories on your computer, so you can use them for different 
projects. The RStudio team has created a number of tools to make this process easier, including
some functionality in the RStudio IDE as well as the `devtools` package [@R-devtools].

## Components of R packages

### R code

Without R code, there's not much point to an R package.^[The exception? A data-only package, 
where the package's purpose is to share data rather than R functions. We'll look at an 
example in the "Collect" section.] The R function code will go in the package's "R"
subdirectory. You can use as many or as few ".R" files in this section as you want to
organize your code. 
Define your functions in this section as you would if you were creating functions in 
an R script. As you become more advanced in writing packages, you may want to start
adapting your function code a bit. For example, the `package::function` notation can 
help keep clear which package you mean for a function to call a function from, and so 
the behavior will be stable regardless of the order that a user loaded packages. 
Also, if you intend for your functions to be used inside tidyverse pipelines, you'll 
want to take a look at the developing [Tidy evaluation](https://tidyeval.tidyverse.org/) book.

Although it is not required for a package you're creating just to use yourself, you 
should think about adding help documentation to your code very early. You can then 
access this documentation through helpfiles in R, just like you do with other R packages.
Even if you are just creating a package for yourself, you should add documentation
for all the "top-level" functions^[**top-level functions.** The functions in your package
that you expect your users (including yourself) to call directly. Your package may also
include some functions that are "helpers", only used within other functions to help 
keep the code clean and avoid replication, but never used directly by package users.]

Years ago, you would need to write this documentation in separate ".Rd" files
and save them in the "man" subdirectory. Now, the easiest thing to do is to use
"roxygen" conventions to write the help information right above each function.
With this style, you'll use a special type of commenting character (`#'`) at the
start of each line with documentation, and there are special codes (e.g.,
`@param` when you're defining one of the function's parameters) to annotate this
code. This help documentation will then, when you build your package, be
converted to the ".Rd" files you used to have to write yourself and saved in the
"man" directory. Several of the resources listed in "Learn more" give more
details on writing and rendering this help documentation.

### DESCRIPTION file

Every package directory must have a DESCRIPTION file, but if you aren't sharing
the package, there's no harm in leaving this as the original template file you
get when you start a package directory using RStudio (you'll try this out in
just a minute). The exception is when your functions rely on functions from
other R packages---in this case, you'll want to add in these dependencies in the
DESCRIPTION file. That way, when you load your package, all of these will also
be loaded.

If you plan to share your package with others, however, and especially if you 
plan to publish your package in CRAN or another repository, you'll need to do 
a lot more with the DESCRIPTION file. This file provides the metadata on your package,
including the authors, version, description, and dependencies on other packages. 
The **R Packages** book referenced in the "Learn More" section has extensive advice
on editing this file for packages meant for publication. 

### Other package components

So far, I've described the barebones elements you'll need to have to create a very 
minimalist package. There are, however, a number of other elements you'll often 
want to add to the package. As with the R code for package functions and the 
DESCRIPTION file, there is a specific location and format required for each of 
these elements within an R package. 

Some of the most common components you'll want to add include example datasets
(which will go in a `data` subdirectory), tutorials for using the package
(`vignettes` subdirectory), and unit tests for the functions (`tests`
subdirectory). To help you set up these extra components, you may want to take a
look at the `use_*` family of functions in the `usethis` [@R-usethis]
package^[The `use_*` family of functions were previously in the `devtools`
package, but they are being deprecated there and moved to the `usethis` package.
The [R Packages](http://r-pkgs.had.co.nz/) book was written before they were
deprecated in `devtools`, and so it discusses them as `devtools` functions. As
long as you install and load the `usethis` package, though, you'll have no
problem following along with the examples in that book.] For example, if you
want to include example data in your package, you should check out the
`use_data` function, while if you want to add unit tests to your package, check
out the `use_test` function. When you run one of these, you'll notice that the
function *adds files and subdirectories* when needed to your R package
directory. This may throw you at first, since most R functions don't make
changes to your file directory, but you'll need those added files to add the
package components, and it's *much* easier to remember a single function call to
run when you need to add one than to remember all the details of which files (and
in what structure) need to be added.

## Try it out

I suggest you try out the following steps to experiment with 
writing your own package. I've included a function for you to build
your package around in the "package.R" file of the R project with 
examples for the workshop. 

- Create a new R Project for your R package
- Build the package
- Move a function into the right place in the package
- Edit the DESCRIPTION file as needed
- Add example data into the right place in the package
- Add some unit tests for the function
- Add documentation using roxygen2 notation
- Add a short vignette

I'll walk you through the first three steps, and then I suggest you use 
Hadley Wickham's **R Packages** book (see the "Learn More" section) to 
get some tips as you try out the rest on your own. 

**Create a new R Project for an R package.** Open RStudio and go to "File" ->
"New Project". Choose "New Directory" and then "R Package". Name your package
"convertr". Then click on "Create Project".

The R Project you've just created will open in RStudio. If you look in the "R"
subdirectory, you'll see there's one file called "hello.R". There's also a 
subdirectory called "man"and three files in the top directory called
".Rbuildignore", "DESCRIPTION", and "NAMESPACE". 

**Build the package.** This project opens with a template---a very small but working package that can
serve as a skeleton as you adapt the files to create your own package. Before you 
do anything to change from the template, try building the package using the 
`build` function from the `devtools` package: 

```{r eval = FALSE}
library(devtools)
build()
```

You should get some output like this: 

```
✔  checking for file ‘/Users/georgianaanderson/Documents/r_workshops/convertr/DESCRIPTION’ ...
─  preparing ‘convertr’:
✔  checking DESCRIPTION meta-information ...
─  checking for LF line-endings in source and make files and shell scripts
─  checking for empty or unneeded directories
─  building ‘convertr_0.1.0.tar.gz’
```

Congratulations! You've built (maybe) your first R package. You should now be able to 
load it like any other R package on computer and run its function. Try running:

```{r eval = FALSE}
library(convertr)
hello()
```

To build the package in a way that you can use it anywhere on your computer, 
go to the "Build" tab in RStudio and click "Install and Restart". As long
as you don't get warning messages, you should be able to load and use the 
package from any R session on your computer. 

**Move a function into the right place in the package.**

Now try adapting the template into your own package. Delete the file
"hello.R" in the "R" subdirectory. Create a new file in this subdirectory
called something like "temp_conversions.R". Copy the code for the
`c_to_f` function, which is in the "package.R" file of the R Project 
with examples for this workshop. Save the file and rebuild the package, 
either with the `build` call to build within that project or with the 
"Install and Restart" button in the "Build" tab to make it available 
anywhere on your computer. 


## Sharing

### When to publish your packages

Once you have some functions that you think others might find useful, you 
should consider creating a package with them and sharing that package with
others. Similarly, if you have a dataset that you think others might 
find useful and that you have permission to share, you should consider creating
a data-only package to use to share that data. 

### Where to publish packages

There are a few ways to publish your R package, and they have different
levels of requirements as well as different levels of recognition (including
for CVs and progress reports). 

First, you can share the "tar.gz" file that's create when you run "Install and Restart"
for an R package from RStudio. The other user can install this file locally using 
`install.packages` with some extra options. For example, if you shared your 
"convertr.tar.gz" package file with someone else, they could install it on their
computer with the call:^[This call assumes that the other person has "convertr.tar.gz"
saved in his or her working directory. Otherwise, use a pathname to direct R to where
the file is saved on the computer.]

```{r eval = FALSE}
install.packages("convertr.tar.gz", repos = NULL, type = "source")
```

Second, and perhaps even easier (certainly for the other user), you can post your
package as a GitHub repository. The other user can then use the `install_github`
function from the `devtools` package to install the package. Unless you set 
the repository to be private, be aware that your package code would be public. 
This means anyone could install and use it, although if you don't publicize it, 
it's unlikely that many people will.

If you want to share your package more widely, you should submit it to a repository. 
The [Comprehensive R Archive Network (CRAN)](https://cran.rstudio.com/) is the most
popular for most scientific fields. The `devtools` package has some tools specifically 
for checking and releasing a package to CRAN. [Bioconductor](https://www.bioconductor.org/)
is the top choice for many bioinformatics packages. 

Finally, an excellent choice is to submit the package to
[ROpenSci](https://ropensci.org/) for peer review. If you are thinking of doing this, 
check ROpenSci's policies on what types of packages are considered "in scope" for it. 
If you submit a package here, it will be openly reviewed by two other R programmers, 
with an open review process. You will then have the chance to revise the package 
code in response to the reviewers' comments. Accepted packages can also be
submitted to CRAN or Bioconductor.

### Extra steps for publishing

If you want to submit your package to CRAN, you'll need to take a few extra
steps to get it ready. These steps aren't necessary for a package you plan to
just use privately, but they wouldn't be a bad idea to try to do even in that
case. For example, you'll need to make sure that the functions in your package
are comprehensively documented. Every function that you expect an end user might
use should have a helpfile. In addition to documenting specific functions, you
should also consider writing overall tutorials (called **vignettes**) that walk
a user through how to use your package in a few example scenarios. You will also
need to make sure that your package can run on different operating systems. If
your package only has straightforward functions written exclusively in R, with
little interaction outside of R on the user's computer, this shouldn't be an
issue. However, you should still check before submitting to CRAN. Several of the
resources listed in "Learn More" have very detailed information on preparing a
package to submit to a public repository.

## Learn more

One wonderful resource for writing and publishing R packages is the book [R
Packages](http://r-pkgs.had.co.nz/) by Hadley Wickham [@wickham2015r]. This book
is available for a reasonable price in paperback (many Barnes & Nobles carry it
in their computer section) and is also available for free online. The author of
this book also created and maintains the **`devtools` package**^[**`devtools`
package.** An R package with functions that facilitate R package development.
These functions are well-documented in the package documentation as well as
through the book [R Packages](http://r-pkgs.had.co.nz/). This package also
includes a function for installing R packages from code posted on GitHub
(`install_github`), which is very useful if you want to find and use packages
that are not yet available on CRAN and other repositories or if you want to use
the latest, "development" version of a package. This package is available on
CRAN and so can be installed with `install.packages("devtools")`.] The "Package
Development" cheatsheet available through [RStudio's
cheatsheets](https://www.rstudio.com/resources/cheatsheets/) complements these
resources.

Along with Roger Peng and Sean Kross, I co-teach a Coursera Specialization on
[Mastering Software Development in
R](https://www.coursera.org/specializations/r), and we also have [an online
book](https://bookdown.org/rdpeng/RProgDA/) on bookdown on the topic. The
official CRAN resource for writing packages, ["Writing R
Extensions"](https://cran.r-project.org/doc/manuals/r-release/R-exts.html), is
available online. However, it gets very technical, and so I recommend starting
with other resources and working up to this document.

To learn more about best practices for creating R, especially for publication,
check out [ROpenSci's handbook](https://ropensci.github.io/dev_guide/). ROpenSci
conducts peer review for R packages, and it's all conducted openly. You can
learn a lot about packages and package development, especially for more complex
packages, by following this peer review package. It's all conducted through [the
"Issues" in a GitHub repo](https://github.com/ropensci/software-review/issues).

If you end up writing and publishing a lot of R packages through your scientific
research, you may find that you need to help your supervisors and peers
understand this research project. One recent article discusses how academic
research in the data science realm is changing, as well as how department chairs
and others can evaluate these new research products, including for promotion and
tenure [@waller2018documenting].